Building Sustainable Software - The CSDMS Approach

### 2015-02-27 hutton_et_al_2014_wssspe.tex
Author additions for JSOR article.
The augmented WSSSPE2 article for the WSSSPE JSOR special issue.
These changes are mostly from Mark, Irina, and Albert.

\csdms{}, The Community Surface Dynamics Modeling System, is an NSF funded
project whose focus is to aid a diverse community of earth and ocean system
model \emph{users} and \emph{developers} to use and create robust software
quickly.  To this end, \csdms{} develops, integrates, archives and disseminates
earth-system models and tools to an international (67 country) community
with the goal of building the set of tools necessary to model the
earth system. Modelers use \csdms{} for access to hundreds of open
source surface-dynamics models and tools, as well as model metadata. Such a
model repository increases model transparency and helps eliminate
duplication by presenting the current state of modeling efforts.
To increase software sustainability, composability and interoperability,
\csdms{} promotes standards that define common modeling interfaces, semantic
mediation between models, and model metadata. Through online resources and
workshops, \csdms{} promotes software engineering best practices, which are
unfamiliar to many developers within our modeling community. For example,
version control, unit testing, continuous integration, test-driven
development, and well-written \emph{clean code} are all topics of the
educational mission of \csdms{}.

@@

The mission of the Community Surface Dynamics Modeling System (\csdms{})
~\cite{peckham2012component} is to help a diverse modeling community toward
common goals and standards. This effort involves creating:
\begin{itemize}
  \item a repository of source code and metadata for open-source models and
    tools (Section~\ref{sec:organizing})
  \item reusable plug-and-play model components and a framework
        within which they can be coupled to create new models
        (Section~\ref{sec:reusable})
  \item an \emph{efficient} and \emph{open} modeling community through
        standards (Section ~\ref{sec:standards}) and education
        (Section~\ref{sec:education})
\end{itemize}

@@

\csdms{} has developed innovative model/component interfaces that promote
model reuse and interoperability,
including the Basic Model Interface (\bmi{} - Section~\ref{sec:bmi}), which
uses
the \csdms{} Standard Names (Section~\ref{sec:standardnames}) and a framework
within which models can be coupled and enhanced with a set of \bmi{}-compatible
service components.
The \csdms{} Web Modeling Tool (\wmt{} - Section~\ref{sec:wmt}) is a web-based application that provides a
graphical interface to this framework and allows users to compose new models
by connecting and configuring components in a simple, browser-based graphical
interface.

@@

Open-source software reduces redundancy, provides transparency through
external review and makes research replicable, which is fundamental to
scientific practices~\cite{ince2012case}.
Software accessibility is a key value of the \csdms{} mission. \csdms{}
ensures code developed by individual researchers or small research teams
remains accessible beyond the lifetime of a specific research project. Code
developers are required to distribute their codes under open-source licenses.

\csdms{} has built an online model repository that not only contains model
source code but also documents the submitted code as model metadata, runs
basic smoke-tests of the models, (optionally) maintains the source code in a
\csdms{}-hosted version control system, and provides easy downloads of the
original source. The repository now contains over 200 models and tools (over
4M lines of code) and has received over 13k model downloads.

@@

Only stable versions of open source
software that are physically hosted through the \csdms{} repository will be
assigned a \doi{}. For model software citations, \csdms{} recommends the
following the DataCite guidelines~\cite{brase2009datacite}.

@@

The \csdms{} Web Modeling Tool\footnote{https://csdms.colorado.edu/wmt}
(\wmt{}) is a web
application that provides an Ajax client-side (the \wmt{}
client\footnote{https://github.com/csdms/wmt-client}) graphical interface
(Figure~\ref{fig:wmt_screenshot}) and a RESTful
server-side database and API (the \wmt{} server\footnote{https://github.com/csdms/wmt}) that allows users to build and run coupled Earth system models on
a high-performance computing cluster from a web browser. \wmt{} was designed
with four objectives:
\begin{itemize}

\item \emph{Accessibility}. As a web-based application, users with internet
      access, have access to \wmt{}.

\item \emph{Integration}. Easily hyperlink from \wmt{} to resources on the 
      \csdms{} modeling portal (model documentation, labs, lectures,
      tutorials)
      %tutorials and movies - or to other resources on the Internet.

@@

%\item \emph{Maintenance}. Because modern browsers tend to adhere to web
%      standards, which lead to fewer cross-compatibility issues than
%      operating systems, only one version of WMT needs to be developed
%      and maintained.

\end{itemize}
With \wmt{}, users can: \emph{run} single component models, build (2-way)
\emph{coupled models}, view and \emph{edit} model parameters, and \emph{share}
saved models with others in the community.

@@

\section{Education}
\label{sec:education}

%\csdms{} constantly evaluates the effectiveness of its tools and standards
%through user feedback from clinics and tutorials and one-on-one interviews
%with \csdms{} members.  Modeling courses at the University of Colorado,
%clinics at the International Summer Institute at the University of Minnesota,
%and clinics at \csdms{} Annual Meetings are the dominant forums for
%this testing and feedback.

@@

%Over the last 4 years over 150 graduate students - from geography, hydrology,
%geology, oceanography, earth sciences and engineering departments - have
%enrolled in intensive \csdms{}-run modeling courses. These courses
%provide an excellent use case for us to test usability and access to our
%tools. Students reported in post-course surveys that they were unfamiliar with
%the use of models on a high performance computing cluster beforehand (rated
%their familiarity as 1.5 average on a Likert scale of 1-5), but ended the
%courses with familiarity and being more comfortable with running and changing
%simulations as well as visualizing results (average rating approaches 4)
%~\cite{overeem2013strategies}.

%Course and clinic materials are posted online as educational resources.
%\csdms{} develops \wmt{}-specific help pages for all models currently
%available as components, so that users can easily learn about model
%parameters and input files. In addition, we are developing web-based video
%tutorials that show how to run models.

@@

\csdms{} provides cyber-infrastructure to promote the quantitative modeling of
earth surface processes. The US National Science Foundation funds \csdms{}, but
the community is international and includes industry and federal agency
representatives in addition to members from academic institutions. The
initiative has been growing with more than 150 people per year. \csdms{} now
fosters a community of approximately 1200 scientists who work on prediction
of the movement of fluids, and the production, erosion, transport, and
deposition of sediment and nutrients in landscapes and seascapes.

>>>>

\csdms{}, The Community Surface Dynamics Modeling System, is an NSF funded
project whose focus is to aid a diverse community of earth and ocean system
model \emph{users} and \emph{developers} to use and create robust software
quickly.  To this end, \csdms{} develops, integrates, archives and disseminates
earth-system models and tools to an international (67 country) community
with the goal of building the set of tools necessary to model
earth surface processes. Modelers use \csdms{} for access to over two hundred open
source surface-dynamics models and tools, as well as model metadata. Such a
model repository increases model transparency and helps eliminate
duplication by presenting the current state of community modeling efforts.
To increase software sustainability, composability and interoperability,
\csdms{} promotes standards that define common modeling interfaces, semantic
mediation between models, and model metadata. Through online resources and
workshops, \csdms{} promotes software engineering best practices, which are
less familiar to many within our community. For example,
version control, unit testing, continuous integration, test-driven
development, and well-written \emph{clean code} are all topics of the
educational mission of \csdms{}.

@@

The mission of the Community Surface Dynamics Modeling System (\csdms{})
~\cite{peckham2012component} is to help move a diverse modeling community toward
common goals and standards. This effort involves creating:
\begin{itemize}
  \item a repository of source code and metadata for open-source models and
    tools (Section~\ref{sec:organizing})
  \item reusable plug-and-play model components and a framework
        within which they can be coupled to create new models
        (Section~\ref{sec:reusable})
  \item an \emph{efficient} and \emph{open} modeling community through
        standards (Section ~\ref{sec:standards}),
        communication and education (Section~\ref{sec:education})
\end{itemize}

@@

\csdms{} has developed innovative model/component interfaces that promote
model reuse and interoperability,
including the Basic Model Interface (\bmi{} - Section~\ref{sec:bmi}), which
uses
the \csdms{} Standard Names (Section~\ref{sec:standardnames}) and a framework
within which models can be coupled and enhanced with a set of \bmi{}-compatible
service components.
The \csdms{} Web Modeling Tool (\wmt{} - Section~\ref{sec:wmt}) is a
web-based application that provides an interface to this
framework.
It allows users to compose new models by connecting and
configuring components in a simple, browser-based graphical interface.
Further, it provides a standardized interface to standalone models;
once a user understands how to run a model in \wmt{},
they can run any model in \wmt{}.

@@

Open-source software reduces redundancy, provides transparency through
external review and makes research replicable, which is fundamental to
scientific practices~\cite{ince2012case}.
Software accessibility is a key value of the \csdms{} mission. \csdms{}
ensures code developed by individual researchers or small research teams
remains accessible beyond the lifetime of a specific research project. Code
developers are required to distribute their software under open-source licenses.

\csdms{} has built an online model 
repository\footnote{http://csdms.colorado.edu/wiki/Model\_download\_portal}
(Figure~\ref{fig:model_repository_screenshot})
where \csdms{} members can contribute code.
The repository stores documentation on submitted code as model metadata,
including a description, version number,
a summary of inputs and outputs,
and author contact information.
Submitted code is given basic smoke tests,
then it is placed under version control in a 
GitHub repository\footnote{https://github.com/csdms-contrib}
(if the contributor already maintains the code on GitHub, 
the code is forked from their repository), 
where it can be forked or cloned by other users.
The repository currently contains over 200 models and tools (over
4M lines of code) and has received over 13k model downloads.

\begin{figure}
  \caption{The \csdms{} Model Repository}
  \begin{center}
    \includegraphics[scale=.25]{csdms_model_repository.eps}
  \end{center}
  \label{fig:model_repository_screenshot}
\end{figure}

@@

Only stable versions of open source
software that are physically hosted through the \csdms{} model repository are
assigned a \doi{}. For model software citations, \csdms{} recommends the
following the DataCite guidelines~\cite{brase2009datacite}.

@@

The \csdms{} Web Modeling Tool\footnote{https://csdms.colorado.edu/wmt}
(\wmt{}) is a web
application that provides an Ajax client-side (the \wmt{}
client\footnote{https://github.com/csdms/wmt-client}) graphical interface
(Figure~\ref{fig:wmt_screenshot}) and a RESTful
server-side database and API (the \wmt{} server\footnote{https://github.com/csdms/wmt})
that allows users to build and run coupled models of earth surface processes on
a high-performance computing cluster from a web browser. \wmt{} was designed
with three objectives:
\begin{itemize}

\item \emph{Accessibility}. As a web-based application, access
      to the Internet implies access to \wmt{}.

\item \emph{Integration}. Easily hyperlink from \wmt{} to resources on the 
      \csdms{} portal (model documentation, labs, lectures,
      tutorials)
      %tutorials and movies - or to other resources on the Internet.

@@

With \wmt{}, users can: \emph{run} single component models, build (2-way)
\emph{coupled models}, view and \emph{edit} model parameters, and \emph{share}
saved models with others in the community.

%% \wmt{}, and the software stack on which it is built, are portable.

@@

\section{Communication and Education}
\label{sec:education}

%\wmt{} provides the ability to run models with an easy to use graphical
%interface and intends to serve novice and more experienced modelers alike.
%The WMT beta version was released in May 2014  as a sequel to the CMT.

\subsection{The CSDMS Portal}

The \csdms{} portal\footnote{http://csdms.colorado.edu}
is the information clearinghouse for the community.
Through the portal,
members can get the latest news,
access model metadata and source code (Section~\ref{sec:organizing}),
view conference and workshop announcements,
view recent job postings,
and access \csdms{} reports, presentations, and publications.
%% including educational presentations and short course
%% materials from \csdms{}-hosted or sponsored workshops.
The portal uses the open-source MediaWiki software,
so \csdms{} members can edit, update and extend existing material,
as well as contribute new material to the portal.

\subsection{User Training}

\csdms{} provides user training in modeling and model coupling
on high-performance computing clusters
through short courses held at 
the University of Colorado,
the International Summer Institute at the University of Minnesota,
and at \csdms{} Annual Meetings.
Course materials are posted to the \csdms{} portal as educational resources.
In addition to training community members,
the clinics also provide an excellent opportunity 
for feedback on using \csdms{} software tools.

\subsection{Programming Courses}

@@

@@

\csdms{} provides cyber-infrastructure to promote the quantitative modeling of
earth surface processes. The US National Science Foundation funds \csdms{}, but
the community is international and includes industry and federal agency
representatives in addition to members from academic institutions. The
initiative has been growing, adding more than 150 people per year. \csdms{} now
fosters a community of over 1200 scientists who work on determining
the movement of fluids, and the production, erosion, transport, and
deposition of sediment and nutrients in landscapes and seascapes.
